\section{Background}
\label{sec:background}

% XXX: should this be a distributed example?
\begin{figure}[t]
\begin{scriptsize}
\begin{lstlisting}
class ShortestPaths
  include Bud

  state do
    table :link, [:from, :to] => [:cost] (*\label{line:spaths-ddl}*)
    scratch :path, [:from, :to, :next_hop, :cost]
    scratch :min_cost, [:from, :to] => [:cost]
  end

  bloom do
    path <= link {|l| [l.from, l.to, l.to, l.cost]} (*\label{line:spaths-proj}*)
    path <= (link*path).pairs(:to => :from) do |l,p| (*\label{line:spaths-join-start}*)
      [l.from, p.to, l.to, l.cost + p.cost]
    end (*\label{line:spaths-join-end}*)

    min_cost <= path.group([:from, :to], min(:cost)) (*\label{line:spaths-group}*)
  end
end
\end{lstlisting}
\end{scriptsize}
\caption{All-pairs shortest paths in Bloom.}
\label{fig:bloom-spaths}
\end{figure}

In this section, we review the Bloom programming language and the CALM program
analysis.  We present a simple program for which the CALM analysis over sets
yields unsatisfactory results.

\subsection{Bloom}
\label{sec:bg-bloom}

Bloom programs are bundles of declarative \emph{statements} about collections of
\emph{facts} (tuples). An \emph{instance} of a Bloom program performs
computation by evaluating its statements over the contents of its local
database. Instances communicate via asynchronous messaging, as described below.

An instance of a Bloom program proceeds through a series of \emph{timesteps},
each containing three phases.\footnote{There is a precise declarative semantics
  for Bloom~\cite{dedalus-semantics,dedalus}, but we describe the language
  operationally for the sake of exposition.} In the first phase, inbound events
(e.g., network messages) are received and represented as facts in
collections. In the second phase, the program's statements are evaluated over
local state to compute all the additional facts that can be derived from the
current collection contents. In some cases (described below), a derived fact is
intended to achieve a ``side effect,'' such as modifying local state or sending
a network message.  These effects are deferred during the second phase of the
timestep; the third phase is devoted to carrying them out.

The initial implementation of Bloom, called \emph{Bud}, allows Bloom logic to be
embedded inside a Ruby program. Figure~\ref{fig:bloom-spaths} shows a Bloom
program represented as an annotated Ruby class. A small amount of imperative
Ruby code is needed to instantiate the Bloom program and begin executing it;
more details are available on the Bloom language website~\cite{bloom}.

\subsubsection{Data model}
\begin{table}[t]
\centering
\begin{tabular}{|l|p{2.32in}|}
\hline
\textbf{Name} & \textbf{Behavior }\\
\hline
\texttt{table} & Persistent storage.\\
\texttt{scratch} & Transient storage.\\
\texttt{channel} & Asynchronous communication. A fact derived into a \texttt{channel} appears in the
database of a remote Bloom instance at a non-deterministic future time.\\
\texttt{periodic} & Interface to the system clock.\\
\texttt{interface} & Interface point between software modules.\\
\hline
\end{tabular}
\caption{Bloom collection types.}
\label{tbl:bloom-collections}
\end{table}

The Bloom data model is based on \emph{collections}.  A collection is an
unordered set of \emph{facts}, akin to a relation in Datalog. The Bud prototype
adopts the Ruby type system rather than inventing its own; hence, a fact in Bud
is just an array of immutable Ruby objects. Each collection has a \emph{schema},
which declares the structure (column names) of the facts in the collection. A
subset of the columns in a collection form its \emph{key}: as in the relational
model, the key columns functionally determine the remaining columns. The
collections used by a Bloom program are declared in a \texttt{state} block. For
example, line~\ref{line:spaths-ddl} of Figure~\ref{fig:bloom-spaths} declares a
collection named \texttt{link} with three columns, two of which form the
collection's key. Ruby is a dynamically typed language, so keys and values in
Bud can hold arbitrary Ruby objects.

Bloom provides five collection types to represent different kinds of state
(Table~\ref{tbl:bloom-collections}). A \texttt{table} stores persistent data: if
a fact appears in a table, it remains in the table in future timesteps (unless it
is explicitly removed). A \texttt{scratch} contains transient data---the content
of scratch collections is emptied at the start of each timestep. Scratches are
akin to SQL views: they are often useful as a way to name intermediate results
or as a ``macro'' construct to enable code reuse. The \texttt{channel}
collection type enables communication between Bloom instances. The schema of a
channel has a distinguished \emph{location specifier} column (prefixed with
``\texttt{@}''); when a fact is derived for a channel collection, it appears in
the database of the Bloom instance at the address given by the location
specifier. The \texttt{periodic} and \texttt{interface} collection types do not
arise in our discussion in this paper; the interested reader is referred to the
Bloom website~\cite{bloom}.

\begin{table}[t]
\centering
\begin{tabular}{|c|l|p{1.85in}|}
\hline
\textbf{Op} & \textbf{Name} & \textbf{Meaning} \\
\hline
\verb|<=| & \emph{merge} & lhs includes the content of rhs in the
current timestep. \\
\hline
\verb|<+| & \emph{deferred merge} & lhs will include the content of rhs in the
next timestep. \\
\hline
\verb|<-| & \emph{deferred delete} & lhs will not include the content of rhs
in the next timestep. \\
\hline
\verb|<~| & \emph{async merge} & (Remote) lhs will include the content of the
rhs at some non-deterministic future timestep. \\
\hline
\end{tabular}
\caption{Bloom operators.}
\label{tbl:bloom-ops}
\end{table}

\ifbool{socc-print-version}{\pagebreak}{}
\subsubsection{Statements}

Each Bloom statement has one or more input collections and a single output
collection.  A statement takes the form: \\ \noindent
\mbox{\hspace{0.25in}\emph{$<$collection-identifier$>$ $<$op$>$
    $<$collection-expression$>$}}\\ \noindent
The left-hand side (lhs) is the name of the output collection and the right-hand
side (rhs) is an expression that produces a collection.  A statement defines how
the input collections are transformed before being included (via set union) in
the output collection. Bloom allows the usual relational operators to be used on
the rhs (selection, projection, join, grouping, aggregation, and negation),
although it adopts a syntax intended to be more familiar to imperative
programmers. In Figure~\ref{fig:bloom-spaths}, line~\ref{line:spaths-proj}
demonstrates projection,
lines~\ref{line:spaths-join-start}--\ref{line:spaths-join-end} perform a join
between \texttt{link} and \texttt{path} using the join predicate
\verb+link.to = path.from+ followed by a projection to four attributes, and
line~\ref{line:spaths-group} shows grouping and aggregation. Bloom statements
appear in one or more \texttt{bloom} blocks, which group together semantically
related statements to aid readability.

% A Bloom program can also include a \texttt{bootstrap} block, which contains
% statements that are evaluated only once when a Bloom instance starts
% executing. \texttt{bootstrap} blocks are typically used for initialization or
% configuration data.

Bloom provides several operators that determine \emph{when} the rhs will be
merged into the lhs (Table~\ref{tbl:bloom-ops}). The \verb|<=| operator performs
standard logical deduction: that is, the lhs and rhs are true at the same
timestep. The \verb|<+| and \verb|<-| operators indicate that facts will be
added to or removed from the lhs collection at the beginning of the \emph{next}
timestep. The \verb+<~+ operator specifies that the rhs will be merged into the
lhs collection at some non-deterministic future time. The lhs of a statement
that uses \verb+<~+ must be a channel; the \verb+<~+ operator captures
asynchronous messaging.

% XXX: does this need to be said?
Bloom allows recursion---i.e., the rhs of a statement can reference the lhs
collection, either directly or indirectly. As in Datalog, certain constraints
must be adopted to ensure that programs with recursive statements have a
sensible interpretation. For deductive statements (\verb+<=+ operator), we
require that programs be \emph{syntactically stratified}~\cite{Apt1988}: cycles
through negation or aggregation are not allowed (unless they contain a deferred
or asynchronous operator)~\cite{dedalus}.

\subsection{CALM analysis}
\label{sec:bg-calm}

Work on deductive databases has long drawn a distinction between
\emph{monotonic} and \emph{non-monotonic} logic programs. Intuitively, a
monotonic program only computes more information over time---it will never
``retract'' a previous conclusion in the face of additional evidence.  In Bloom
(and Datalog), a simple conservative test for monotonicity is based on program
syntax: selection, projection, and join are monotonic, while aggregation and
negation are not.

The CALM theorem connects the theory of monotonic logic with the practical
problem of distributed consistency~\cite{Alvaro2011,Hellerstein2010}.  All
monotonic programs are ``eventually consistent'' or \emph{confluent}: for any
given input, all program executions result in the same final state regardless of
network non-determinism~\cite{Ameloot2011,dedalus-confluence}.  Hence, monotonic
logic is a useful building block for loosely consistent distributed programming.

According to the CALM theorem, distributed inconsistency may only occur at
\emph{points of order}: program locations where the output of an asynchronously
derived value is consumed by a non-monotonic operator~\cite{Alvaro2011}.  This
is because asynchronous messaging results in non-deterministic arrival order,
and non-monotonic operators may produce different conclusions when evaluated
over different subsets of their inputs.  For example, consider a Bloom program
consisting of a pair of collections $A$ and $B$ (both fed by asynchronous
channels) and a rule that sends a message whenever an element of $A$ arrives
that is not in $B$.  This program is non-monotonic and exhibits non-confluent
behavior: the messages sent by the program will depend on the order in which the
elements of $A$ and $B$ arrive.

We have implemented a conservative static program analysis in Bloom that follows
directly from the CALM theorem.  Programs that are free from non-monotonic
constructs are ``blessed'' as confluent: producing the same output on different
runs or converging to the same state on multiple distributed replicas.
Otherwise, programs are flagged as potentially inconsistent.  To achieve
consistency, the programmer either needs to rewrite their program to avoid the
use of non-monotonicity or introduce a coordination protocol to ensure that a
consistent ordering is agreed upon at each of the program's points of order.
Coordination protocols incur additional latency and reduce availability in the
event of network partitions, so in this paper we focus on coordination-free
designs---that is, monotonic programs.

\subsubsection{Limitations of set monotonicity}
The original formulation of the CALM theorem considered only programs that
compute more facts over time---that is, programs whose \emph{sets} grow
monotonically. Many distributed protocols make progress over time, but their
notion of ``progress'' is often difficult to represent as a growing set of
facts. For example, consider the Bloom program in
Figure~\ref{fig:bloom-nm-quorum}. This program receives votes from one or more
clients (not shown) via the \texttt{vote\_chn} channel. Once at least
\texttt{QUORUM\_SIZE} votes have been received, a message is sent to a remote
node to indicate that quorum has been reached
(line~\ref{line:bloom-quorum-msg}). This program resembles a ``quorum vote''
subroutine that might be used by an implementation of Paxos~\cite{Lamport1998}
or quorum replication~\cite{Gifford1979}.

Intuitively, this program makes progress in a semantically monotonic fashion:
the set of received votes grows and the size of the \texttt{votes} collection
can only increase, so once a quorum has been reached it will never be
retracted. Unfortunately, the current CALM analysis would regard this program as
non-monotonic because it contains a point of order: the grouping operation on
line~\ref{line:bloom-nm-quorum}.

To solve this problem, we need to introduce a notion of program values that
``grow'' according to a partial order other than set containment. We do this by
extending Bloom to operate over arbitrary lattices rather than just the
set lattice.

%  We present a
% complete language in the following section, but the intuition can be observed in
% Figure~\ref{fig:lattice-quorum}. Votes are accumulated into a set lattice
% (line~\ref{line:quorum-set-accum}), but the size of the set is represented as an
% \texttt{lmax} lattice (line~\ref{line:quorum-lmax}): that is, a number that
% never decreases. Hence, a threshold test ``$\ge k$'' on an \texttt{lmax} lattice
% is monotonic map onto the boolean lattice: that is, the \texttt{quorum\_done}
% predicate goes from false to true (and then remains true).

\begin{figure}[t]
\begin{scriptsize}
\begin{lstlisting}
QUORUM_SIZE = 5
RESULT_ADDR = "example.org"

class QuorumVote
  include Bud

  state do
    channel :vote_chn, [:@addr, :voter_id]
    channel :result_chn, [:@addr]
    table   :votes, [:voter_id]
    scratch :cnt, [] => [:cnt]
  end

  bloom do
    votes      <= vote_chn {|v| [v.voter_id]}
    cnt        <= votes.group(nil, count(:voter_id)) (*\label{line:bloom-nm-quorum}*)
    result_chn <~ cnt {|c| [RESULT_ADDR] if c >= QUORUM_SIZE} (*\label{line:bloom-quorum-msg}*)
  end
end
\end{lstlisting}
\end{scriptsize}
\caption{A non-monotonic Bloom program that waits for a quorum of votes to be received.}
\label{fig:bloom-nm-quorum}
\end{figure}
