\section{Adding Lattices to Bloom}
\label{sec:lang}

This section introduces \lang, an extension to Bloom that allows monotonic
programs to be written using arbitrary lattices. We begin by reviewing the
algebraic properties of lattices used in CRDTs and note the applicability of
monotone functions and morphisms in that context. We then introduce the basic
concepts of \lang and detail the built-in lattices provided by the language. We
also show how users can define their own lattice types.

% is this the right place for this?
When designing \lang, we decided to extend Bloom to include support for lattices
rather than building a new language from scratch. Hence, \lang is backward
compatible with Bloom and was implemented with relatively minor changes to the
Bud runtime. We describe how code written using lattices can interoperate with
traditional Bloom collections in Section~\ref{sec:bloom-interop}.

\subsection{Definitions}
\label{sec:lattice-defn}
A \emph{bounded join semilattice} is a triple $\langle S, \sqcup, \bot\rangle$,
where $S$ is a set, $\sqcup$ is a binary operator
(called ``join'' or ``least upper bound''), and $\bot \in S$. The operator
$\sqcup$ is associative, commutative, and idempotent. The $\sqcup$ operator induces a partial order $\leq_S$ on the elements of $S$: $x \leq_S y$ if $x \sqcup y = y$.
% 
% For all $x, y \in S$, $x
% \sqcup y = z$, where $x \leq_S z, y \leq_S z$, and there is no $z' \ne z \in S$
% such that $z' \leq_S z$ (where $\leq_S$ is the partial order of poset $S$). 
Note
that although $\leq_S$ is only a partial order, the least upper bound is defined for
all elements $x,y \in S$. The distinguished element $\bot$ is the smallest
element in $S$:  $x \sqcup \bot = x$ for every $x \in S$. For brevity,
we use the term ``lattice'' to mean ``bounded join semilattice'' in the rest of
this paper. We use the informal term ``merge function'' to mean ``least upper
bound.''

A \emph{monotone function} from poset $S$ to poset $T$ is a function $f: S \to
T$ such that $\forall a,b \in S: a \leq_S b \Rightarrow f(a) \leq_T f(b)$. That
is, $f$ maps elements of $S$ to elements of $T$ in a manner that respects the
partial orders of both posets.

% XXX: mention that morphisms must be distributive with respect to the lub of
% their domain, whereas monotone functions don't need to be?
A \emph{morphism} from lattice $\langle X, \sqcup_X, \bot_X\rangle$ to lattice
$\langle Y, \sqcup_Y, \bot_Y\rangle$ is a function $g: X \to Y$ such that
$g(\bot_X) = \bot_Y$ and $\forall a,b \in X: g(a \sqcup_X b) = g(a) \sqcup_Y
g(b)$. Intuitively, $g$ maps elements of $X$ to elements of $Y$ in a way that
preserves the lattice properties.  Note that morphisms are monotone functions
but the converse is not true in general.

\subsection{Language concepts}
\begin{figure}[t]
\begin{scriptsize}
\begin{lstlisting}
QUORUM_SIZE = 5
RESULT_ADDR = "example.org"

class QuorumVoteL
  include Bud

  state do
    channel :vote_chn, [:@addr, :voter_id]
    channel :result_chn, [:@addr]
    lset    :votes (*\label{line:quorum-lset-decl}*)
    lmax    :cnt
    lbool   :quorum_done
  end

  bloom do
    votes       <= vote_chn {|v| v.voter_id} (*\label{line:quorum-set-accum}*)
    cnt         <= votes.size (*\label{line:quorum-size}*)
    quorum_done <= cnt.gt_eq(QUORUM_SIZE) (*\label{line:quorum-threshold}*)
    result_chn  <~ quorum_done.when_true { [RESULT_ADDR] } (*\label{line:quorum-when-true}*)
  end
end
\end{lstlisting}
\end{scriptsize}
\caption{A monotonic \lang program that waits for a quorum of votes to be received.}
\label{fig:lattice-quorum}
\end{figure}

\lang allows both lattices and collections to represent state. A lattice is
analogous to a collection type in Bloom, while a \emph{lattice element}
corresponds to a particular collection. For example, the \texttt{lset} lattice
is similar to the \texttt{table} collection type provided by Bloom; an element
of the \texttt{lset} lattice is a particular set. In the terminology of
object-oriented programming, a lattice is a class that obeys a certain interface
and an element of a lattice is an instance of that
class. Figure~\ref{fig:lattice-quorum} contains an example \lang program.

As with collections, the lattices used by a \lang program are declared in a
\texttt{state} block. More precisely, a state block declaration introduces an
identifier that is associated with a lattice element; over time, the binding
between identifiers and lattice elements is updated to reflect state changes in
the program. For example, line~\ref{line:quorum-lset-decl} of
Figure~\ref{fig:lattice-quorum} declares an identifier \texttt{votes} that is
mapped to an element of the \texttt{lset} lattice. As more votes are received,
the lattice element associated with the \texttt{votes} identifier changes (it
moves ``upward'' in the \texttt{lset} lattice). When a lattice identifier is
declared, it is initially bound to the value $\bot$, the smallest element in the
lattice. For example, an \texttt{lset} lattice initially contains the empty set.

\begin{table*}[t]\small
\begin{center}
\begin{tabular}{|l|p{0.88in}|l|l|p{1.44in}|p{1.00in}|}
\hline
\textbf{Name} & \textbf{Description} & \textbf{Least element ($\bot$)} & \textbf{Merge($a,b$)} & \textbf{Morphisms} &
\textbf{Monotone functions}\\
\hline
\hline
\texttt{lbool} & Boolean lattice \newline (false $\to$ true) & \texttt{false} & $a \lor
b$ & \texttt{when\_true(\&blk)} $\to$ \texttt{v} & \\
\hline
\texttt{lmax} & Max over an \newline ordered domain & $-\infty$ & $max(a, b)$ &
\texttt{gt(n)} $\to$ \texttt{lbool}\newline
\texttt{gt\_eq(n)} $\to$ \texttt{lbool}\newline
\texttt{$\mathtt{+}$(n)} $\to$ \texttt{lmax}\newline
\texttt{$\mathtt{-}{}$(n)} $\to$ \texttt{lmax} & \\
\hline
\texttt{lmin} & Min over an \newline ordered domain & $\infty$ & $min(a, b)$ &
\texttt{lt(n)} $\to$ \texttt{lbool}\newline
\texttt{lt\_eq(n)} $\to$ \texttt{lbool}\newline
\texttt{$\mathtt{+}$(n)} $\to$ \texttt{lmin}\newline
\texttt{$\mathtt{-}{}$(n)} $\to$ \texttt{lmin} & \\
\hline
\texttt{lset} & Set of values & empty set & $a \cup b$ &
\texttt{intersect(lset)} $\to$ \texttt{lset}\newline
\texttt{project(\&blk)} $\to$ \texttt{lset}\newline
\texttt{product(lset)} $\to$ \texttt{lset}\newline
\texttt{contains?(v)} $\to$ \texttt{lbool}
& \texttt{size()} $\to$ \texttt{lmax}\\
\hline
\texttt{lpset} & Set of non-\newline{}negative numbers & empty set & $a \cup b$ &
\texttt{intersect(lpset)} $\to$ \texttt{lpset}\newline
\texttt{project(\&blk)} $\to$ \texttt{lpset}\newline
\texttt{product(lpset)} $\to$ \texttt{lpset}\newline
\texttt{contains?(v)} $\to$ \texttt{lbool}
& \texttt{size()} $\to$ \texttt{lmax}\newline
\texttt{sum()} $\to$ \texttt{lmax} \\
\hline
\texttt{lbag} & Multiset of values & empty multiset & $a \cup b$ &
\texttt{intersect(lbag)} $\to$ \texttt{lbag}\newline
\texttt{project(\&blk)} $\to$ \texttt{lbag}\newline
\texttt{card(v)} $\to$ \texttt{lmax}\newline
\texttt{contains?(v)} $\to$ \texttt{lbool}\newline
\texttt{$\mathtt{+}$(lbag)} $\to$ \texttt{lbag}
& \texttt{size()} $\to$ \texttt{lmax}\\
\hline
\texttt{lmap} & Map from keys to \newline{}lattice values & empty map & see text&
\texttt{intersect(lmap)} $\to$ \texttt{lmap}\newline
\texttt{project(\&blk)} $\to$ \texttt{lmap}\newline
\texttt{key\_set()} $\to$ \texttt{lset}\newline
\texttt{at(v)} $\to$ \texttt{any-lattice}\newline
\texttt{key?(v)} $\to$ \texttt{lbool}
& \texttt{size()} $\to$ \texttt{lmax}\\
\hline
\end{tabular}
\caption{Built-in lattices in \lang. Note that \texttt{v} denotes a Ruby value,
  \texttt{n} denotes a number, and \texttt{blk} indicates a Ruby code block
  (anonymous function).}
\label{tbl:built-in-lattices}
\end{center}
\end{table*}


\subsubsection{Statements in \lang}
Statements take the same form in both Bloom and \lang: \\ \noindent
\mbox{\hspace{0.25in}\emph{$<$identifier$>$ $<$op$>$
    $<$expression$>$}}\\ \noindent
The identifier on the lhs can refer to either a set-oriented collection or a
lattice element. The expression on the rhs can contain both traditional
relational operators (applied to Bloom collections) and methods invoked on
lattices.  Lattice methods are similar to methods in an object-oriented language
and are invoked using the standard Ruby method invocation syntax. For example,
line~\ref{line:quorum-size} of Figure~\ref{fig:lattice-quorum} invokes the
\texttt{size} method on an element of the \texttt{lset} lattice.

If the lhs is a lattice, the statement's operator must be either \verb|<=| or
\verb|<+| (instantaneous or deferred deduction, respectively). The meaning of
these operators is that, at either the current or the following timestep, the lhs
identifier will take on the result of applying the lattice's least upper bound
to the lhs and rhs lattice elements. The intuition remains the same as in Bloom:
the rhs value is ``merged into'' the lhs lattice, except that the semantics of
the merge operation are defined by the lattice's least upper bound operator. We
require that the lhs and rhs refer to a lattice of the same type.

\lang does not support deletion (\verb|<-| operator) for lattices. Lattices do
not directly support asynchronous communication (via the \verb|<~| operator) but
lattice elements can be embedded into facts that appear in channels
(Section~\ref{sec:lattice-embedding}).

\subsubsection{Lattice methods}
\lang statements compute values over lattices by invoking methods on lattice
elements. Just as a subset of the relational algebra is monotonic, some lattice
methods are monotone functions (as defined in Section~\ref{sec:lattice-defn}). A
monotone lattice method guarantees that, if the lattice on which the method is
invoked grows (according to the lattice's partial order), the value returned by
the method will grow (according to the return value's lattice type). For
example, the \texttt{size} method provided by the \texttt{lset} lattice is
monotone because as more elements are added to the set, the size of the set
increases. From a CRDT perspective, a lattice's monotone methods constitute a
``safe'' interface of operations that can be invoked in a distributed setting
without risk of inconsistency.

A lattice method's signature indicates its monotonicity properties. \lang
distinguishes between methods that are monotone and a subset of monotone methods
that are \emph{morphisms}. Section~\ref{sec:lattice-defn} defines the properties
that a morphism must satisfy, but the intuition is that a morphism on lattice
$T$ can be distributed over $T$'s least upper bound. For example, the
\texttt{size} method of the \texttt{lset} lattice is not a morphism. To see why,
consider two elements of the \texttt{lset} lattice, $\{1,2\}$ and $\{2,3\}$.
The \texttt{size} method is not a morphism because $size(\{1,2\}
\sqcup_{\mathtt{lset}} \{2,3\}) \ne size(\{1,2\}) \sqcup_{\mathtt{lmax}}
size(\{2,3\})$. Morphisms can be evaluated more efficiently than monotone
methods, as we discuss in Section~\ref{sec:lattice-eval-strat}.

Lattices can also define non-monotonic methods. Using a non-monotonic lattice
method is analogous to using a non-monotonic relational operator in Bloom: the
Bud interpreter stratifies the program to ensure that the input value is
computed to completion before allowing the non-monotonic method to be
invoked. \lang encourages developers to minimize the use of non-monotonic
constructs: as the CALM analysis suggests, non-monotonic reasoning may need to
be augmented with coordination to ensure consistent results.
% Stratification implies that programs with cycles through non-monotonic
% lattice methods will be rejected. 

Every lattice defines a non-monotonic \texttt{reveal} method that returns a
representation of the lattice element as a plain Ruby value. For example, the
\texttt{reveal} method on an \texttt{lmax} lattice returns a Ruby integer. This
method is non-monotonic because once the underlying Ruby value has been
extracted from the lattice, \lang cannot ensure that subsequent code uses the
value in a monotonic fashion.

\subsection{Built-in lattices}
\label{sec:lattice-built-ins}

Table~\ref{tbl:built-in-lattices} lists the lattices included with \lang. The
built-in lattices provide support for several common notions of ``progress'': a
predicate that moves from false to true (\texttt{lbool}), a numeric value that
strictly increases or strictly decreases (\texttt{lmax} and \texttt{lmin},
respectively), and various kinds of collections that grow over time
(\texttt{lset}, \texttt{lpset}, \texttt{lbag}, and \texttt{lmap}). The behavior
of most of the lattice methods should be unsurprising, so we do not describe
every method in this section.

The \texttt{lbool} lattice represents conditions that, once satisfied, remain
satisfied. For example, the \texttt{gt} morphism on the \texttt{lmax} lattice
takes a numeric argument $n$ and returns an \texttt{lbool}; once the
\texttt{lmax} exceeds $n$, it will remain $>n$. The \texttt{when\_true} morphism
takes a Ruby block; if the \texttt{lbool} element has the value \texttt{true},
\texttt{when\_true} returns the result of evaluating the block. For example, see
line~\ref{line:quorum-when-true} in
Figure~\ref{fig:lattice-quorum}. \texttt{when\_true} is similar to an ``if''
statement.\footnote{Observe that an ``else'' clause would test for an upper
  bound on the final lattice value, which is a non-monotonic property!}

The collection-like lattices support familiar operations such as union,
intersection, and testing for the presence of an element in the collection. The
\texttt{project} morphism takes a code block and forms a new collection by
applying the code block to each element of the input collection. Elements for
which the code block returns \texttt{nil} are omitted from the output
collection, which allows \texttt{project} to be used as a filter.

The \texttt{lmap} lattice associates keys with values. Keys are immutable Ruby
objects and values are lattice elements. For example, a web application could
use an \texttt{lmap} to associate session IDs with an \texttt{lset} containing
the pages visited by that session. The \texttt{lmap} merge function takes the
union of the key sets of its input maps. If a key occurs in both inputs, the two
corresponding values are merged using the appropriate lattice merge
function. Note that the \texttt{at(v)} morphism returns the lattice element
associated with key \texttt{v} (or $\bot$ if the \texttt{lmap} does not contain
\texttt{v}).

The \texttt{lpset} lattice is an example of how \lang can be used to encode
domain-specific knowledge about an application. If the developer knows that a
set will only contain non-negative numbers, the sum of those numbers increases
monotonically as the set grows. Hence, \texttt{sum} is a monotone function of
\texttt{lpset}; in contrast, taking the sum of the elements of an \texttt{lset}
is non-monotonic in general.

\subsection{User-defined lattices}
\begin{figure}[t]
\begin{scriptsize}
\begin{lstlisting}[deletekeywords={lset}]
class Bud::SetLattice < Bud::Lattice
  wrapper_name :lset (*\label{line:lset-wrapper-name}*)

  def initialize(x=[])
    @v = x.uniq # Remove duplicates from input
  end

  def merge(i)
    self.class.new(@v | i.reveal)
  end

  morph :intersect do |i| (*\label{line:lset-morph-begin}*)
    self.class.new(@v & i.reveal) (*\label{line:lset-reveal}*)
  end (*\label{line:lset-morph-end}*)

  morph :contains? do |i|
    Bud::BoolLattice.new(@v.member? i)
  end

  monotone :size do (*\label{line:lset-monotone-begin}*)
    Bud::MaxLattice.new(@v.size)
  end (*\label{line:lset-monotone-end}*)
end
\end{lstlisting}
\end{scriptsize}
\caption{Example implementation of the \texttt{lset} lattice.}
\label{fig:lattice-lset}
\end{figure}

\label{sec:lattice-api}
The built-in lattices are sufficient to express many programs. However, \lang
also allows developers to create custom lattices to capture domain-specific
behavior. To define a new lattice, a developer creates a Ruby class that meets a
certain API contract. Figure~\ref{fig:lattice-lset} shows a simple
implementation of the \texttt{lset} lattice using a Ruby array for
storage.\footnote{We omit a few \texttt{lset} methods for brevity. While we use
  an array for simplicity, this is inefficient (e.g., eliminating duplicates is
  expensive). The built-in \texttt{lset} uses a hash-based set data type.}

A lattice class must inherit from the built-in \texttt{Bud::Lattice} class and
must also define two methods:
\begin{compactitem}
\item \texttt{initialize(i)}: given a Ruby object $i$, this method constructs a
  new lattice element that ``wraps'' $i$ (this is the standard Ruby syntax for
  defining a constructor). If $i$ is \texttt{nil} (the null reference), this
  method returns $\bot$, the least element of the lattice.

\item \texttt{merge(e)}: given a lattice element $e$, this method returns the
  least upper bound of \emph{self} and \emph{e}. This method must satisfy the
  algebraic properties of least upper bound as summarized in
  Section~\ref{sec:lattice-defn}---in particular, it must be commutative,
  associative, and idempotent. Note that \emph{e} and \emph{self} must be
  instances of the same class.
\end{compactitem}
Lattices can also define any number of monotone functions, morphisms, and
non-monotonic methods. The syntax for declaring morphisms and monotone functions
can be seen in lines~\ref{line:lset-morph-begin}--\ref{line:lset-morph-end} and
\ref{line:lset-monotone-begin}--\ref{line:lset-monotone-end} of
Figure~\ref{fig:lattice-lset}, respectively. Note that lattice elements are
\emph{immutable}---that is, lattice methods (including \texttt{merge} methods)
must return new values, rather than destructively modifying any of their inputs.
% explain why?

%  (e.g., \texttt{merge} functions construct a new lattice element
% rather than destructively modifying one of their inputs). Efficient lattice
% implementations will often \emph{share structure} on merge operations, as is
% common practice for immutable data structures in functional programming
% languages~\cite{Okasaki1999}. % XXX: not the right place for this?


% Because lattice methods can contain arbitrary Ruby code, the lattice developer
% should be careful to ensure that lattice methods satisfy the appropriate
% algebraic properties. For example, implementing a lattice method might require
% examining the underlying Ruby value contained in another lattice element. This
% can be done using the \texttt{reveal} method (e.g.,
% line~\ref{line:lset-reveal} in Figure~\ref{fig:lattice-lset}).  Since
% \texttt{reveal} is not itself monotonic, developers should use it carefully
% when implementing monotonic lattice methods.  Tweak this text

% Because Ruby is dynamically typed, there will often be constraints on the legal
% inputs to lattice methods that cannot be enforced until runtime. For example,
% initializing an \texttt{lbool} lattice with a non-boolean value is not
% supported. The developer of a lattice implementation should check for invalid
% inputs and signal errors by raising a Ruby exception.

% Lattice elements are \emph{immutable} (e.g., \texttt{merge} functions construct
% a new lattice element rather than destructively modifying one of their
% inputs). Efficient lattice implementations will often \emph{share structure} on merge
% operations, as is common practice for immutable data structures in functional
% programming languages~\cite{Okasaki1999}. % XXX: not the right place for this?

% XXX: awkward placement for this paragraph
Custom lattices must define a keyword that can be used in \lang state
blocks. This is done using the \texttt{wrapper\_name} class method. For example,
line~\ref{line:lset-wrapper-name} of Figure~\ref{fig:lattice-lset} means that
``\texttt{lset :foo}'' in a state block will introduce an identifier
\texttt{foo} that is associated with an instance of \texttt{Bud::SetLattice}.

\subsection{Integration with set-oriented logic}
\label{sec:bloom-interop}
\lang provides two features to ease integration of lattice-based code with Bloom
rules that use set-oriented collections.

\subsubsection{Converting collections into lattices}
% XXX: this ignores the fact that Bloom collections consist of sets of tuples,
% whereas implicit fold works for sets of singleton values
This feature enables an intuitive syntax for merging the contents of a
set-oriented collection into a lattice. If a statement has a Bloom collection on
the rhs and a lattice on the lhs, the collection is converted into a lattice
element by ``folding'' the lattice's merge function over the collection. That
is, each element of the collection is converted to a lattice element (by
invoking the lattice constructor) and then the resulting lattice elements are
merged together via repeated application of the lattice's \texttt{merge}
method. In our experience, this is usually the behavior intended by the user.

For example, line~\ref{line:quorum-set-accum} of Figure~\ref{fig:lattice-quorum}
contains a Bloom collection on the rhs and an \texttt{lset} lattice on the
lhs. This statement is implemented by constructing a singleton \texttt{lset} for
each fact in the rhs collection and then merging the sets together. The
resulting \texttt{lset} is then merged with the \texttt{votes} lattice
referenced by the lhs.

\subsubsection{Collections with embedded lattice values}
\label{sec:lattice-embedding}
\lang allows lattice elements to be used as columns of facts in Bloom
collections. This feature allows \lang programs to use a mixture of Bloom-style
relational operators and lattice method invocations, depending on which is more
convenient. Bloom also provides several collection types with special semantics
(e.g., network communication, durable storage); allowing lattice elements to be
embedded into collections avoids the need to create a redundant set of
facilities for lattices.

Consider a simple \lang statement that derives facts with an embedded lattice
element as a column:
\begin{verbatim}
    t1 <= t2 {|t| [t.x, cnt]}
\end{verbatim}
where \texttt{t1} and \texttt{t2} are Bloom collections, \texttt{cnt} is a
lattice, and the key of \texttt{t1} is its first column. Note that \texttt{cnt}
might change over the course of a single timestep (specifically, \texttt{cnt}
can increase according to the lattice's partial order). This might result in
multiple \texttt{t1} facts that differ only in the second column, which would
violate \texttt{t1}'s key.

To resolve this situation, \lang allows multiple facts to be derived that differ
only in their embedded lattice values; those facts are merged into a single fact
using the lattice's merge function. This is similar to specifying a procedure
for how to resolve key constraint violations, a feature supported by some
databases~\cite{oracle-conflict,sqlite-on-conflict}. For similar reasons,
lattice elements cannot be used as keys in Bloom collections.%  It might be
% possible to relax this restriction in certain ``safe'' cases, but we have not
% found this limitation to be problematic to date.

% Storing lattice elements as columns of tuples in set-oriented collections
% introduces several challenges. Consider a simple \lang statement that derives
% tuples with a lattice element as a column:
% \begin{verbatim}
%     t1 <= t2 {|t| [t.x, cnt]}
% \end{verbatim}
% where \texttt{t1} and \texttt{t2} are Bloom collections, \texttt{cnt} is a
% lattice, and the first column of \texttt{t1} is the collection's key. The value
% associated with \texttt{cnt} can change over the course of a single timestep
% (\texttt{cnt} can move ``upward'' according to the lattice's partial
% order). This might result in multiple \texttt{t1} tuples with different values
% for the second column, which would violate \texttt{t1}'s key (the first column
% of \texttt{t1} would not functionally determine a single value for the second
% column).

% This problem could be avoided by placing constraints on the evaluation order of
% statements: for example, we could require that all potential changes to
% \texttt{cnt} be completed before statements that embed \texttt{cnt} could be
% evaluated. This would effectively stratify the program according to lattice
% embedding statements and disallow cycles through lattice
% embeddings~\cite{Apt1988}.  This would reduce the expressiveness of the
% language; it also seems unsatisfying to require stratification of monotonic
% programs.

% Instead, \lang allows multiple facts to be derived that differ only in their
% embedded lattice values; those facts are merged into a single fact using the
% lattice's merge function. This is similar to specifying a procedure for how to
% resolve key constraint violations, a feature supported by some
% databases~\cite{oracle-conflict,sqlite-on-conflict}. Note that the safety of the
% merge procedure is guaranteed by the lattice properties: the final value is
% deterministic and is the least upper bound of all the merged
% values. Nevertheless, we felt this behavior should be manifest in the program
% syntax because tuples in Datalog relations are traditionally immutable. Hence,
% this ``lattice merge'' behavior is only enabled for columns that are prefixed
% with ``\texttt{!}'' in collection declarations. If the
% annotation is not specified, lattice elements embedded in Bloom collections
% cannot change; it is up to the programmer to avoid key constraint violations.

\subsection{Confluence in \lang}
\label{sec:calmL}

We now describe how the notion of \emph{confluence} (invariance to message
reordering) can be generalized from Bloom to \lang programs. In recent work, we
provided a model-theoretic characterization of confluence for programs written
in Dedalus, the formal language on which Bloom is based~\cite{dedalus-confluence}. 
These results apply directly to Bloom, whose semantics are grounded in those of
Dedalus.  The result of a distributed computation performed in Bloom may be
viewed as the set of \emph{ultimate models} or eventual states of the program,
given a fixed input and sufficient time for messages to be delivered.  If a
program has exactly one ultimate model for every input, we say that it is
confluent: all message delivery orders result in the same eventual state.

To reason about confluence in \lang, we first observe that lattices are
guaranteed to have inflationary behavior over time: a lattice value only
increases.  Hence if we include lattices in the output of an otherwise confluent
\lang program, they will not increase the number of ultimate models.  It is not
enough, however, that this local property of lattice objects holds: we must also
show that anything a \lang program \emph{does} with a lattice value is
monotonic.  Monotone lattice functions provide a mechanism for reasoning about
composition of lattices or between lattices and collections.  By including
monotone lattice functions among the ``safe,'' monotonic constructs provided by
the Bloom language, we can easily extend our CALM analysis to \lang.
