%\documentclass[10pt,twocolumn]{article}
\documentclass{sig-alternate}
\usepackage{color,graphicx}
\usepackage{hhline}
\usepackage{alltt}
\usepackage{url}
\usepackage{times}
\usepackage{mathptmx}
\usepackage{xspace}
\usepackage{cite}
\usepackage{color}
\usepackage{epsfig}
\usepackage{theorem}

\input{boon}

\newcommand{\comment}{\textcolor{red}}
\newcommand{\link}[1]{{\small \url{#1}}}
\newcommand{\note}[1]{}

\def\Pitu{P2\xspace}
\def\Sys{\Pitu}
\def\sysx{\Pitu}
\def\Dlog{{\em NDlog}\xspace}
\def\link{\texttt{\#link}\xspace}
\def\bilink{\texttt{\#bilink}\xspace}
\def\dblspace{\xspace\xspace}
\def\eg{{\sc $e.g.,~$}}
\def\ie{{\sc $i.e.,~$}}
\def\Eg{{\sc $E.g.,~$}}
\def\Ie{{\sc $I.e.,~$}}
\def\manyquads{\quad\quad\quad\quad\quad\quad\quad\quad}

%\setlength{\oddsidemargin}{-0in}
%\setlength{\evensidemargin}{-0in}
%\setlength{\hoffset}{-0.25in}
%\setlength{\textwidth}{7in}
%\setlength{\topmargin}{0.1in}
%\setlength{\voffset}{-0.25in}
%\setlength{\headheight}{-0.3pt}
%\setlength{\headsep}{0pt}
%\setlength{\textheight}{9.3in}

\addtolength{\parindent}{-1mm}
\addtolength{\parskip}{-0.2mm}
%\addtolength{\belowcaptionskip}{-2mm}
%\addtolength{\abovecaptionskip}{-2mm}

\addtolength{\dblfloatsep}{-2mm}
\addtolength{\dbltextfloatsep}{-2mm}
\setlength{\columnsep}{1.45pc}
\linespread{0.943}
%\linespread{1}

% Compact itemize and enumerate. Note that they use the same counters and
% symbols as the usual itemize and enumerate environments.
\def\compactify{\itemsep=0pt \topsep=0pt \partopsep=0pt \parsep=0pt}
 \let\latexusecounter=\usecounter
 \newenvironment{CompactItemize}
   {\def\usecounter{\compactify\latexusecounter}
    \begin{itemize}}
   {\end{itemize}\let\usecounter=\latexusecounter}
 \newenvironment{CompactEnumerate}
   {\def\usecounter{\compactify\latexusecounter}
    \begin{enumerate}}
   {\end{enumerate}\let\usecounter=\latexusecounter}


\begin{document}

% SIGMOD Copyright
%\conferenceinfo{SIGMOD 2006}{June 27-29, 2006, Chicago, Illinois, USA}
%\CopyrightYear{2006}
%\crdata{1-59593-256-9/06/0006}

% Avoid SIGMOD Copyright
%\def\copyrightspace{}
\toappear{Appears in the ACM SIGMOD International Conference on Management of Data, Chicago, IL, USA, June 2006.}  


\newcommand{\dataloglabel}[1]{\mbox{\bf #1:}\hfil}
\newenvironment{datalog}
  {\begin{list}{}%
      {\it \small \renewcommand{\makelabel}{\dataloglabel}%
    \setlength{\parsep}{-2pt}%
    }%
  }%
{\end{list}}

\newcommand{\datalogspace}{\textcolor[gray]{1}{.}\hspace{0.8in}}


\renewcommand{\ttdefault}{cmtt}
\newcounter{foo1}
\newenvironment{myenumerate}{\begin{list}{\arabic{foo1}.}{
      \usecounter{foo1}
      \setlength{\itemsep}{0mm}
      \setlength{\topsep}{0mm}
      \setlength{\itemsep}{-2pt}
}}{ \end{list}}

\newenvironment{mylist}{\begin{list}{$\bullet$}
    {
      \setlength{\itemsep}{0mm}
      \setlength{\topsep}{0mm}}
    }
{\end{list}}


\newcommand{\calS}{\mathcal{S}}
\newcommand{\calD}{\mathcal{D}}
\newcommand{\eat}[1]{}



% \title{Declarative Networking with\\Distributed Recursive Query
% Processing}
\title{Declarative Networking:\\Language, Execution and Optimization}

\author{Boon Thau Loo\thanks{\scriptsize UC Berkeley authors funded
  by NSF grants 0205647, 0209108, and 0225660, and a
  gift from Microsoft.} \dblspace Tyson Condie$^*$ \dblspace Minos Garofalakis$^\dagger$\dblspace David E.\ Gay$^\dagger$\dblspace Joseph
  M.\ Hellerstein$^*$\dblspace\\Petros Maniatis$^\dagger$\dblspace Raghu Ramakrishnan$^\ddagger$\dblspace Timothy
  Roscoe$^\dagger$\dblspace Ion Stoica$^*$\\ {\small $^*$UC Berkeley, $^\dagger$Intel Research Berkeley and
  $^\ddagger$University of Wisconsin-Madison}}

\date{}

\maketitle

\begin{abstract} 
  The networking and distributed systems communities have recently
  explored a variety of new network architectures, both for
  application-level overlay networks, and as prototypes for a
  next-generation Internet architecture.  In this context, we have
  investigated {\em declarative networking}: the use of a distributed
  recursive query engine as a powerful vehicle for accelerating
  innovation in network
  architectures~\cite{declareRoute,declareOverlays,singhEurosys}. 
  Declarative networking represents a significant new application area for
  database research on recursive query processing. In this paper, we
  address fundamental database issues in this domain. First, we
  motivate and formally define the Network Datalog (\Dlog) language
  for declarative network specifications.
%   We introduce the concept of {\em link-restricted}
%   rules, which can be syntactically guaranteed to be executable via
%   single-node derivations and message passing on an underlying network
%   graph. 
  Second, we introduce and prove correct relaxed versions of the
  traditional semi-\naive query evaluation technique, to overcome fundamental
  problems of the traditional technique in an asynchronous distributed
  setting.  Third, we consider the dynamics of network state, and
  formalize the ``eventual consistency'' of our programs even when
  bursts of updates can arrive in the midst of query execution.
  Fourth, we present a number of query optimization opportunities that
  arise in the declarative networking context, including applications
  of traditional techniques as well as new optimizations. Last, we
  present evaluation results of the above ideas implemented in our
  \Sys declarative networking system, running on 100 machines over the
  Emulab network testbed.
\end{abstract}



\input{intro}
\input{queryModel}
\input{queryPro}
\input{dynamic}
\input{queryOpt}
\input{eval}
\input{conclude}

%\linespread{0.95}
{\scriptsize
\bibliographystyle{abbrv}
\bibliography{biblio}
}
\input{proofs}
\end{document}
