\documentclass{sigplanconf}

\usepackage[usenames, dvipsnames]{color}
\usepackage{times}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{wrapfig}
\usepackage{url}
\usepackage{amsmath, amssymb, amsthm}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{comment}
\usepackage{alltt}
\usepackage{appendix}
\usepackage{algorithm}
\usepackage{algorithmic}

\pdfinfo{/Title (Dedalus: Datalog in Time and Space)}

\usepackage{txfonts}
\newcommand{\Tau}{\mathcal{T}}
\newcommand{\SDedalus}{\mathcal{S}}
\newcommand{\Consts}{\mathcal{C}}
\newcommand{\Vars}{\mathcal{A}}
%\newcommand{\pos}{\protect{$_{pos}$}}
%\newcommand{\nega}{\protect{$_{neg}$}}
% RCS: Would like to use the above ones, but can't get them to work in Dedalus env.
\newcommand{\pos}{\_pos}
\newcommand{\nega}{\_neg}

\newcommand{\jmh}[1]{{\textcolor{red}{#1 -- jmh}}}
\newcommand{\paa}[1]{{\textcolor{blue}{#1 -- paa}}}
\newcommand{\rcs}[1]{{\textcolor{green}{#1 -- rcs}}}
\newcommand{\nrc}[1]{{\textcolor{magenta}{#1 -- nrc}}}
\newcommand{\wrm}[1]{{\color{BurntOrange}{#1 -- wrm}}}
\newcommand{\smallurl}[1]{{\small \url{#1}}}

\newtheorem{example}{Example}
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{observation}{Observation}

%\def\slang{synchronous Dedalus\xspace}
\def\lang{\textsc{Dedalus}\xspace}
\def\slang{\textsc{Dedalus\ensuremath{_{{0}}}}\xspace}
%\def\synclang{{Dedalus\ensuremath{_{\large 0}}}\xspace}
\newcommand{\naive}      {na\"{\i}ve\xspace}
\newcommand{\Naive}      {Na\"{\i}ve\xspace}
%dedalus environment for code

\newenvironment{Dedalus}{
\vspace{0.5em}\begin{minipage}{0.95\textwidth}%\linespread{1.3}
\begin{alltt}\fontsize{9pt}{9pt}\selectfont}
{\end{alltt}\end{minipage}\vspace{0.5em}}

\newcommand{\dedalus}[1]{\texttt{\fontsize{9pt}{9pt}\selectfont #1}}


\begin{document}

\conferenceinfo{POPL '10}{Austin, TX, USA} 
\copyrightyear{2011} 
\copyrightdata{[to be supplied]}

\title{{\huge{\bf\lang}}:
Datalog in Time and Space} 
%%Format\titlenote{(Produces the permission block, copyright information and page numbering). For use with ACM\_PROC\_ARTICLE-SP.CLS V2.6SP. Supported by ACM.}}
%
% You need the command \numberofauthors to handle the "boxing"
% and alignment of the authors under the title, and to add
% a section for authors number 4 through n.

%\numberofauthors{6}

\authorinfo{Peter Alvaro \and William R. Marczak \and Neil Conway \and Joseph M. Hellerstein}
           {University of California, Berkeley}
           {\{paa, wrm, nrc, jmh\}@eecs.berkeley.edu}
\authorinfo{David Maier}
           {Portland State University}
           {maier@cs.pdx.edu}
\authorinfo{Russell Sears}
           {Yahoo! Research}
           {dunno@rustys.email}



\maketitle

\begin{abstract} 
%
Recent research has explored using Datalog-based
languages to express a distributed system as a set of logical
invariants~\cite{boom-techr,loo-sigmod06}.  Two properties of distributed
systems proved difficult to model in Datalog.  First, the state of any such system
evolves with its execution.  Second, deductions in these systems may be
arbitrarily delayed, dropped, or reordered by the unreliable network links
they must traverse.  Previous efforts addressed the former by extending
Datalog to include updates, key constraints, persistence and events, and
the latter by assuming ordered and reliable delivery while ignoring delay.
These details have a semantics outside Datalog, which increases the complexity of
the language or its interpretation, and forces programmers to
think operationally.  We argue that the missing component from these previous
languages is a notion of {\em time}.

In this paper we present {\bf \lang}, a foundation language for programming and
reasoning about distributed systems.  \lang reduces to a subset of
Datalog~\cite{ullmanbook} with negation, aggregate functions, successor \wrm{is
``successor" specific enough? should we say ``infinite dense order''} and
choice \wrm{cite?}.  \lang also admits an explicit representation of logical
time.
%
We show that \lang provides a declarative foundation for the two signature
features of distributed systems: mutable state, and asynchronous processing and
communication.  
%%Given these two features, we address three important properties
%%of programs in a domain-specific manner: a notion of {\em safety} appropriate
%%to non-terminating computations, {\em stratified} monotonic reasoning with
%%negation over time, and efficient evaluation over time via a simple execution
%%strategy.  
Given these features, we illustrate how powerful verification techniques from the
Datalog literature may be preserved or reinterpreted in the new domain, including 
conservative syntactic checks for a notion of {\em safety} appropriate to 
non-terminating computation, and {\em stratified} monotonic reasoning with negation 
over time.
%%We also provide conservative syntactic checks for our temporal
%%notions of safety and stratification.  
We also explore the use of similar static analysis techniques to verify properties unique 
to systems with asynchronous communication, including commutativity and idempotence
of in-flight messages.
Our experience implementing
full-featured systems in variants of Datalog suggests that \lang is well-suited
to the specification of rich distributed services and protocols, and provides
both cleaner semantics and richer tests of correctness.

\end{abstract}

\input{intro}
\input{motivation}
\input{foundation}

\input{constructs}
\input{safety}
%%\input{proofs}
\input{update}
\input{async}
%%\input{broadcast}
\input{lang}
\input{properties}
\input{casestudy}
\input{relwork}
\input{future}
\input{acks}
\bibliographystyle{abbrvnat}
\bibliography{popl,declarativity}

%%\appendix
\input{appendix}

%%\input{attic}

\end{document}
