\subsection{The Bloom Vision}
In our experience~\cite{boom-eurosys, netdb}, much of the intellectual
complexity in a typical distributed system resides in \emph{state
  management}: for example, managing protocol and session state, replicating and
  partitioning that state, and restoring system state to consistency after
  failures. 
We advocate a {\em data-centric\/} approach to the specification and implementation 
of such systems, in which all state is represented uniformly as data and programs
are expressed as continuous declarative queries.
By allowing the programmer to focus on applying transformations to
data, much of the boilerplate code required in conventional languages can be
eliminated. Instead, the programmer can focus on reasoning about what is
difficult.

The widespread popularity of MapReduce~\cite{mapreduce-osdi} is a prominent
example of the promise of data-centric programming. MapReduce programmers focus
on applying transformations to sets of data: the MapReduce framework handles the
messy details of partitioning, distribution and fault tolerance. MapReduce
raises the level of abstraction from the traditional Van Neumann model to a
functional dataflow model.

While MapReduce is limited to batch processing, it points the way toward a
promising approach to building distributed systems. In this project, we propose
to investigate whether data-centric techniques can be generalized and used to
build a broad range of distributed systems. To achieve this goal, we are
designing \emph{Bloom}, a new declarative language for distributed computing,
and a stack of associated software tools.
