%\documentclass{acm_proc_article-sp-sigmod09}
\documentclass{sig-alternate}
%\documentclass{sig-alt-full}

\usepackage[usenames, dvipsnames]{color}
\usepackage{times}
%%\usepackage{graphicx}
%%\usepackage{boxedminipage}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{wrapfig}
\usepackage{url}
%%\usepackage{verbatim}
%%\usepackage{latexsym}
\usepackage{amsmath, amssymb}
%\usepackage{amsthm}

\usepackage{alltt}
\usepackage{appendix}
\usepackage{algorithm}
\usepackage{algorithmic}

\usepackage{txfonts}
\newcommand{\Tau}{\mathcal{T}}
\newcommand{\SDedalus}{\mathcal{S}}
\newcommand{\Consts}{\mathcal{C}}
\newcommand{\Vars}{\mathcal{A}}
%\newcommand{\pos}{\protect{$_{pos}$}}
%\newcommand{\nega}{\protect{$_{neg}$}}
% RCS: Would like to use the above ones, but can't get them to work in Dedalus env.
\newcommand{\pos}{\_pos}
\newcommand{\nega}{\_neg}

\newcommand{\jmh}[1]{{\textcolor{red}{#1 -- jmh}}}
\newcommand{\paa}[1]{{\textcolor{blue}{#1 -- paa}}}
\newcommand{\rcs}[1]{{\textcolor{green}{#1 -- rcs}}}
\newcommand{\nrc}[1]{{\textcolor{magenta}{#1 -- nrc}}}
\newcommand{\wrm}[1]{{\color{BurntOrange}{#1 -- wrm}}}
\newcommand{\smallurl}[1]{{\small \url{#1}}}
\newcommand{\etal}       {{\em et al.\/}}

\newtheorem{example}{Example}
\newdef{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{observation}{Observation}

%\def\slang{synchronous Dedalus\xspace}
\def\lang{\textsc{Dedalus}\xspace}
\def\slang{\textsc{Dedalus\ensuremath{_{{0}}}}\xspace}
%\def\synclang{{Dedalus\ensuremath{_{\large 0}}}\xspace}
\newcommand{\naive}      {na\"{\i}ve\xspace}
\newcommand{\Naive}      {Na\"{\i}ve\xspace}
%dedalus environment for code

\newenvironment{Dedalus}{
\vspace{0.5em}\begin{minipage}{0.95\textwidth}%\linespread{1.3}
\begin{alltt}\fontsize{9pt}{9pt}\selectfont}
{\end{alltt}\end{minipage}\vspace{0.5em}}

\newcommand{\dedalus}[1]{\texttt{\fontsize{9pt}{9pt}\selectfont #1}}


\begin{document}

\conferenceinfo{ACM PODS}{'10 Indianapolis, IN, USA}
\title{{\huge{\bf\lang}}:
Datalog in Time and Space} 
%%Format\titlenote{(Produces the permission block, copyright information and page numbering). For use with ACM\_PROC\_ARTICLE-SP.CLS V2.6SP. Supported by ACM.}}
%
% You need the command \numberofauthors to handle the "boxing"
% and alignment of the authors under the title, and to add
% a section for authors number 4 through n.

\numberofauthors{6}

\author{
%
Peter Alvaro$^\ast$ \quad William R. Marczak$^\ast$ \quad Neil
Conway$^\ast$  \\
Joseph M. Hellerstein$^\ast$ \quad 
David Maier$^\dagger$ \quad Russell Sears$^\ast$ 
\\\\
%
\fontsize{10}{10}\selectfont\itshape 
%\vspace{0.05in}
$^\ast$University of California, Berkeley \quad $^\dagger$Portland State
University\\\\ \fontsize{9}{9}\selectfont\ttfamily\upshape
%
\{palvaro,wrm,nrc,hellerstein,sears\}@cs.berkeley.edu \quad maier@cs.pdx.edu
%
}

\maketitle

\begin{abstract} 
%
Recent research has explored using Datalog-based
languages to express a distributed system as a set of logical
invariants~\cite{boom-techr,loo-sigmod06}.  Two properties of distributed
systems proved difficult to model in Datalog.  First, the state of any such system
evolves with its execution.  Second, deductions in these systems may be
arbitrarily delayed, dropped, or reordered by the unreliable network links
they must traverse.  Previous efforts addressed the former by extending the
Datalog model to admit updates, key constraints, persistence and events, and
the latter by assuming ordered and reliable delivery while ignoring delay.
These details have a semantics outside Datalog, which increases the complexity of
the language or its interpretation, and forces programmers to
think operationally.  We argue that the missing component from these previous
languages is a notion of {\em time}.  In this paper we present {\bf \lang}, a
foundation language for programming and reasoning about distributed systems.
\lang reduces to a subset of Datalog~\cite{ullmanbook} with negation, aggregate
functions, successor and choice, and admits an explicit representation of time into the
logic language.  
%Thus, \lang allows concepts such as persistence, updateable state, key constraints, and message delay and loss to be described in a purely declarative fashion.  
We show that \lang provides a declarative foundation for the two signature features of 
distributed systems: mutable state, and asynchronous processing and communication.
Given these two features, we address three important properties of programs in 
a domain-specific manner: a notion of {\em safety} appropriate to non-terminating computations, 
{\em stratified} monotonic reasoning with negation over time, and efficient evaluation
over time via a simple execution strategy.  We also provide conservative syntactic checks
for our temporal notions of safety and stratification.  Our experience implementing 
full-featured systems in variants of Datalog suggests that \lang is well-suited to the specification of
rich distributed services and protocols, and provides both cleaner semantics and richer tests of correctness.
%\lang extends Datalog with a notion of time --- every \lang relation has an
%integer attribute that intuitively represents when a given tuple is true.  We
%propose a sugared syntax in which rule head annotations indicate whether
%their conclusions hold at the same time as their premises, in the immediate
%next timestep, or at some unknown timestep.
% We further introduce the notion of \emph{temporal stratification}, a
% syntactically checkable property of \lang programs similar to stratification,
% and show that the class of temporally stratifiable programs in \lang
% is a subset of the class of modularly stratified~\cite{ross-syntactic,modular}
% programs in Datalog.
%
\end{abstract}

\input{intro}
\input{foundation}

\input{constructs}
\input{safety}
\input{proofs}
\input{update}
\input{async}
\input{broadcast}
\input{relwork}
\input{future}
\bibliographystyle{abbrv}
\bibliography{pods,declarativity}

%%\appendix
%%\input{appendix}

%%\input{attic}

\end{document}
