
\subsection{Stratification in {\large{\bf\slang}}}
\label{sec:strat}
We first turn our attention to the semantics of programs with negation.  As we
will see, the inclusion of time enables a syntactic stratification condition
for programs, and the existence of a unique model that corresponds to
intuition~\cite{local-strat}.






\begin{lemma} \label{lemma:no-neg-unique}
A \slang program without negation 
has a unique minimal model.
\end{lemma}

\begin{proof} 
A \slang program without negation 
is a pure Datalog
program.  Every pure Datalog program has a unique minimal model. 
\end{proof}



We define syntactic stratification of a \slang program the same way it is
defined for a Datalog program:

\begin{definition}
A \slang program is \emph{syntactically stratifiable} if there
exists no cycle with a negative edge 
in the program's
predicate dependency graph.
\end{definition}


We may evaluate such a program in {\em stratum order} as described in the
Datalog literature~\cite{ullmanbook}.
It is easy to see that any syntactically stratified \slang instance has a
unique perfect model~\cite{local-strat} because it is a syntactically stratified Datalog program.




However, many programs we are interested in expressing are not syntactically
stratifiable.  Fortunately, we are able to define a syntactically checkable
notion of {\em temporal stratifiability} of \slang programs that maps to a
subset of locally stratifiable Datalog programs.








\begin{definition} 
The \emph{deductive reduction} of a \slang program $P$ is
the subset of $P$ consisting of exactly the deductive rules in $P$.
\end{definition}

\begin{definition} 
A \slang program is \emph{temporally stratifiable} if its deductive
reduction is syntactically stratifiable.
\end{definition}

\begin{lemma}
\label{lemma:temp-strat-uniq}
Any temporally stratifiable \slang instance $P$ has a unique perfect model.
\end{lemma} 

\begin{proof}
Every temporally stratifiable \slang instance is locally
stratifiable~\cite{local-strat}, and thus has a unique perfect model.



\end{proof}


\begin{example}
A simple temporally stratifiable \slang program that is not syntactically stratifiable.

\begin{Dedalus}
persist[p\pos, p\nega, 3]  
  
p_pos(A, B, T) \(\leftarrow\)
  insert\_p(A, B, T);

p_neg(A, B, T) \(\leftarrow\)
  p_pos(A, B, T),
  delete\_p(T);
\end{Dedalus}

In the \slang program above, \emph{insert\_p} and \emph{delete\_p} are captured
in EDB relations.  This reasonable program is unstratifiable because $p\pos \succ
p\nega \land p\nega \succ p\pos$.  But because the successor relation is
constrained such that $\forall A,B, successor(A, B) \rightarrow B > A$, any
such program is locally stratified on time suffixes.  Therefore, we have
$p\pos_{n} \not\succ^+ p\_neg_{n} \not\succ^+ p\pos_{n+1}$; informally, earlier values
do not depend on later values.
\end{example}




