\section{Related Work}


\subsection{Deductive Databases and Updateable State}

Many deductive database systems, including LDL~\cite{ldl} and Glue-Nail~\cite{glue-nail}, admit procedural semantics to deal with updates using an
assignment primitive.  In contrast, languages proposed by Cleary and Liu~\cite{harmful,deductiveupdates,starlog} retain a purely logical 
interpretation by admitting temporal extensions into their syntax and interpreting assignment or update as a composite operation
across timesteps~\cite{deductiveupdates} rather than as a primitive.  We follow the approach of Datalog/UT~\cite{deductiveupdates} in that we use explicit time suffixes to enforce a stratification condition,
but differ in several significant ways.
First, we model persistence explicitly in our language, so that like updates, it is specified as a composite operation across timesteps.
Partly as a result of this, we are able to enforce stricter constraints on the allowable time suffixes in rules: a program may only specify what deductions are visible
in the current timestep, the immediate next timestep, and \emph{some} future timestep, as opposed to the free use of intervals allowed in rules in Liu et al.  


U-Datalog~\cite{udatalog-neg} addresses updates using syntax annotations that establish different interpretations for the set of updated 
relations and the IDB, interpreting update atoms as constraints and using constraint logic programming techniques to test for inconsistent
derivations.  Similarly, Timed Concurrent Constraint Programming (TCCP)~\cite{tdccp,tccp} handles nonmonotonic constructs in a CLP framework
by outputting a new (possibly diminished) store and constraint program at each timestep.  

Our temporal approach to representing state change most closely resembles the
Statelog language~\cite{statelog}.  
By contrast, our motivation is
the logical specification and implementation of distributed systems, and our 
principal contribution is the use of time to model both local state change and communication
over unreliable networks.

Lamport's TLA+~\cite{tla} is a language for specifying concurrent systems in
terms of constraints over valuations of state and temporal logic that describes
admissible transitions.  Two distinguishing features of \lang with respect to
TLA+ are our minimalist use of temporal constructs (\texttt{next} and
\texttt{async}), and our unified treatment of temporal and other attributes of
facts; this enables the full literature of Datalog to be applied to both
temporal and instantaneous properties of programs.

\subsection{Distributed Systems}

Significant recent work
(\cite{boom-eurosys,Belaramani:2009,Chu:2007,Loo2009-CACM}, etc.) has focused on
applying deductive database languages to the problem of specifying and
implementing network protocols and distributed systems.  Implementing
distributed systems entails a data store that changes over time, so any useful
implementation of such a language addresses the updateable state issue in some
manner.  Existing distributed deductive languages such as NDlog and Overlog
adopt a \emph{chain of fixpoints} interpretation. Evaluation proceeds in three
phases:

\begin{enumerate}
\item Input from the external world, including network messages, clock interrupts and host language calls, is collected.
\item Time is frozen, the union of the local store and the batch of events is taken as EDB, and the program is run to fixpoint.
\item The deductions that cause side effects (e.g., deletions, updates, network messages and host language callbacks) are dealt with.  
\end{enumerate}

Unfortunately, the language descriptions give no careful specification of how and when deletions and updates
should be made visible, so the third step is a ``black box.''  Loo et al.~\cite{Loo2009-CACM} proved that classes of programs with certain 
monotonicity properties (i.e., programs without negation or fact deletion)
are equivalent (specifically, eventually consistent) when evaluated globally (via a single fixpoint computation) or in a distributed setting in which the 
\emph{chain of fixpoints} interpretation is applied at each participating node, and no messages are lost.
Navarro et al.~\cite{navarro} proposed an alternate syntax that addressed key ambiguities in Overlog, including the
\emph{event creation vs.\ effect} ambiguity.  Their solution solves the problem by introducing procedural semantics to the interpretation of 
the augmented Overlog programs.  A similar analysis was offered by Mao~\cite{Mao2009}.


