\section{Case Study: Causal Delivery}
\label{sec:causal}

In the previous section, we showed how \lang increases the space of programs
that can be shown to be confluent. However, many distributed protocols are not
\emph{intended} to have a single deterministic outcome. Instead, these protocols allow timing conditions to drive the
choice of one among several acceptable outcomes. For example, in a vector clock
system, the assignment of clock values to messages differs depending on
message ordering. Nevertheless, the system does not exhibit arbitrary
non-determinism: for example, the clock at each node should move upward over time.

In this section, we show how \lang can be used to implement two non-confluent
protocols: vector clocks and causal delivery. Although we cannot guarantee a
global correctness property like confluence, lattices can still provide strong local
guarantees. For example, we can ensure that vector clock entries increase monotonically,
which agrees with our intuitive notion that these protocols make progress over
time. These guarantees are similar to the interface guarantees provided by
CRDTs, with the notable difference that they are not confined to individual modules: monotone functions allow safe mappings
between lattice values.

As a secondary benefit, we observe that our programs are concise and natural.  In the case of vector clocks,
our code is \textasciitilde{}20$\times$ shorter than a comparable Java program, and exhibits a
direct visual correspondence to Wikipedia's protocol pseudocode. This
increases our confidence in \lang as a language that is well suited to common distributed programming tasks.

% Vector clocks are a classical mechanism for recording the causal relationships
% between events in a distributed system~\cite{Fidge1988,Mattern1989}. In this
% section, we first show how vector clocks can be implemented in a monotonic
% fashion using \lang. We then use \lang to implement a classical algorithm for
% point-to-point causal message delivery~\cite{Schiper1989}. The implementation of
% both protocols in \lang is concise and readable---this suggests that \lang is
% suitable for common distributed programming tasks. Perhaps more significantly,
% both protocols are monotonic. This agrees with our intuition that these
% protocols make ``progress'' over time, and hence gives us more confidence in the
% correctness of our designs.

% Concede that confluence is not an appropriate correctness criteria

\subsection{Vector clocks}
\begin{figure}[t]
\begin{scriptsize}
\begin{lstlisting}
module VectorClock
  state do
    lmap :my_vc
    lmap :next_vc
    scratch :in_msg, [:addr, :payload] => [:clock] (*\label{line:ddl-scratch-start}*)
    scratch :out_msg, [:addr, :payload]
    scratch :out_msg_vc, [:addr, :payload] => [:!clock] (*\label{line:ddl-lattice-merge}*)(*\label{line:ddl-scratch-end}*)
  end

  bootstrap do
    my_vc <= {ip_port => Bud::MaxLattice.new(0)} (*\label{line:vc-init-clock}*)
  end

  bloom do
    next_vc <= my_vc
    next_vc <= out_msg { {ip_port => my_vc.at(ip_port) + 1} } (*\label{line:vc-out-increment}*)
    next_vc <= in_msg  { {ip_port => my_vc.at(ip_port) + 1} } (*\label{line:vc-in-increment}*)
    next_vc <= in_msg  {|m| m.clock} (*\label{line:vc-in-merge}*)
    my_vc <+ next_vc

    out_msg_vc <= out_msg {|m| [m.addr, m.payload, next_vc]} (*\label{line:vc-out-stamp}*)
  end
end
\end{lstlisting}
\end{scriptsize}
\caption{Vector clocks in \lang.}
\label{fig:vector-clock-src}
\end{figure}

\emph{Vector clocks} are a classical mechanism for recording the causal
relationships between events in a distributed
system~\cite{Fidge1988,Mattern1989}.  A vector clock is an array containing a
logical clock for each node in the distributed system. Each node keeps a local
vector clock that it updates when it sends and receives messages; the node's
vector clock reflects how up-to-date its local knowledge is with respect to the
other nodes in the system.

To quote directly from Wikipedia, a vector clock observes the following rules for clock updates:
\emph{
\begin{compactitem}
\item Initially all clocks are zero.
\item Each time a process experiences an internal event, it increments its own logical clock in the vector by one.
\item Each time a process prepares to send a message, it increments its own logical clock in the vector by one and then sends its entire vector along with the message being sent.
\item Each time a process receives a message, it increments its own logical clock in the vector by one and updates each element in its vector by taking the maximum of the value in its own vector clock and the value in the vector in the received message (for every element).''~\cite{vcwikipedia}
\end{compactitem}}
\jmh{The correspondence isn't perfect.  But the Wikipedia echo would be awesome if we can do it.}

% In a system that uses vector clocks, each node $n$ initializes its local vector
% clock $v[n] = 0$. A node updates $v$ by following three rules when sending and
% receiving events:
% \begin{compactenum}
% \item
%   Before sending a message, $n$ increments $v[n]$ and includes $v$ in the
%   outgoing message.
% \item
%   Upon receipt of a message, $n$ increments $v[n]$.
% \item
%   Upon receipt of a message, $n$ merges the vector clock in the message with its
%   own vector clock by taking the element-wise max of the logical clock
%   values.\footnote{If a clock value for a node occurs in one vector clock but
%     not the other, the missing clock value is assumed to be $0$.}
% \end{compactenum}
Figure~\ref{fig:vector-clock-src} contains a complete implementation of vector
clocks in \lang. A vector clock is represented as a map lattice that associates
node IDs with \texttt{lmax} values. Each \texttt{lmax} represents a single
logical clock---naturally, a clock can only increase over time.
\texttt{ip\_port} returns the IP address and port number of the current \lang
instance, which is used as a node ID. Incoming messages appear in the
\texttt{in\_msg} collection. Messages that are intended to be sent begin as
tuples in \texttt{out\_msg}; once a message has been stamped with the local
node's vector clock, it appears in \texttt{out\_msg\_vc}.

The key logic appears in lines~\ref{line:vc-init-clock}, \ref{line:vc-out-increment},
\ref{line:vc-in-increment}, and \ref{line:vc-in-merge}: these four statements
are a direct translation of the four rules \jmh{not quite!} given in the Wikipedia pseudocode. Line~\ref{line:vc-init-clock} initializes each node's local vector clock.
Line~\ref{line:vc-out-stamp} stamps outgoing messages with the updated
vector clock value.  Note that if multiple outgoing messages are sent in the
same \lang timestep, they will contain the same embedded vector clock (and
\texttt{my\_vc} will only be incremented once). This is reasonable, since those
messages are concurrent.

We argue that Figure~\ref{fig:vector-clock-src} is about as readable as the pseudocode from Wikipedia, requiring
only six \lang statements. This compares favorably with implementations of
vector clocks in traditional programming languages. For example, the
\texttt{VectorClock} class included with the Voldemort key-value store consists
of 216 lines of Java source code, not including whitespace or
comments~\cite{voldemort-vector-clock}.

As noted above, we would not expect an implementation of vector clocks to be
confluent; accordingly, Figure~\ref{fig:vector-clock-src} is not a monotonic
\lang program because of the use of \texttt{scratch} collections
(lines~\ref{line:ddl-scratch-start}--\ref{line:ddl-scratch-end}). Scratch
collections are not monotonic because they are emptied at the start of each
timestamp. Although the vector clock program is not confluent, the statements in
the program only invoke monotone lattice methods. Hence, \texttt{my\_vc}
increases monotonically, and \texttt{next\_vc} is greater than or equal to both
\texttt{my\_vc} and all the messages received in the current timestep. These program properties formally guarantee our intuition about how vector clocks change over time.

Note that because the statements in Figure~\ref{fig:vector-clock-src} involve
only monotonic operations, the statements can be executed in any order. The
``lattice embedding'' feature described in Section~\ref{sec:lattice-embedding}
ensures that the \texttt{out\_msg\_vc} tuple contains a single vector clock that
reflects all the messages received in the current timestep.

%  Vector clocks can also
% be implemented in Bloom but they require the use of non-monotonic operators
% (e.g., incrementing a node's clock requires non-monotonic state update using
% $\verb|<+|$ and $\verb|<-|$, and merging the clock values in inbound messages
% requires a \texttt{max} aggregate).

\subsection{Causal delivery}
\begin{figure}[t]
\begin{scriptsize}
\begin{lstlisting}
module CausalDelivery
  include VectorClock (*\label{line:causal-d-import}*)

  state do
    channel :chn, [:@dst, :payload] => [:clock, :ord_buf]
    table :recv_buf, chn.schema
    lmap :ord_buf
  end

  bloom :outbound_msg do
    chn <~ out_msg {|m| [m.addr, m.payload, next_vc, ord_buf]} (*\label{line:out-msg-embed}*)
    ord_buf <+ out_msg {|m| {m.addr => next_vc} } (*\label{line:ord-buf-out-msg}*)
  end

  bloom :inbound_msg do
    recv_buf <= chn
    in_msg <= recv_buf do |m| (*\label{line:deliver-msg-start}*)
      m.ord_buf.at(ip_port).lt_eq(my_vc).when_true { (*\label{line:deliver-msg-lt-eq}*)
        [m.addr, m.payload, m.clock]
      }
    end (*\label{line:deliver-msg-end}*)
    ord_buf <+ in_msg {|m| m.ord_buf} (*\label{line:ord-buf-in-msg}*)
  end
end
\end{lstlisting}
\end{scriptsize}
\caption{Point-to-point causal delivery in \lang. Note that \texttt{in\_msg} and
\texttt{out\_msg} are defined by the \texttt{VectorClock} module.}
\label{fig:causal-delivery-src}
\end{figure}

Next, we use \lang to build a more complex distributed protocol that makes use
of vector clocks. A \emph{causal delivery} protocol ensures that messages are
delivered in an order that is consistent with the ``happens before'' relation
between events~\cite{Lamport1978}. Causal delivery is a common
building block for distributed protocols.

Figure~\ref{fig:causal-delivery-src} contains a \lang program that implements a
classical algorithm for point-to-point (unicast) causal delivery proposed by
Schiper et al.~\cite{Schiper1989}. Note the Ruby \texttt{import} statement on
line~\ref{line:causal-d-import}, which loads the \lang state declarations and
statements from the \texttt{VectorClock} module.

In the Schiper et al.\ protocol, each node records its knowledge about the state
of the vector clocks at every other node. In
Figure~\ref{fig:causal-delivery-src}, this knowledge is kept in
\texttt{ord\_buf}, an \texttt{lmap} that associates node IDs with vector
clocks. A node's \texttt{ord\_buf} is updated when messages are sent or
delivered (lines~\ref{line:ord-buf-out-msg} and \ref{line:ord-buf-in-msg},
respectively). A node also embeds both its vector clock and \texttt{ord\_buf} in
outgoing messages (line~\ref{line:out-msg-embed}). Finally, the protocol ensures
that a node will delay delivering a message $m$ until it can be sure that it
will never receive a message that causally predeces $m$. This is done by
comparing the knowledge the sender had about the recipient's vector clock with
the recipient's current vector clock; once the recipient's vector clock passes
the sender's knowledge, the message can be safely delivered
(lines~\ref{line:deliver-msg-start}--\ref{line:deliver-msg-end}, respectively).

Note that checking whether a message has been delivered requires testing whether
one vector clock (in the \texttt{ord\_buf} embedded in the message) is smaller
than the node's vector clock (line~\ref{line:deliver-msg-lt-eq}). In general,
this is not monotonic: the truth value of an inequality between two monotonically increasing counters can oscillate over time.  In this case, however, the message's \texttt{ord\_buf} vector clock is \emph{immutable}, because it is
embedded as a column of a collection that does not allow ``lattice merges''. So
the inequality predicate can be determined to be monotonic.

Because causal delivery builds upon vector clocks, it is also not
confluent. However, because the statements in
Figure~\ref{fig:causal-delivery-src} only invoke monotone methods, we can see
that a node's \texttt{ord\_buf} strictly increases over time.
