\documentclass[10pt]{sigplanconf}
\usepackage{color}
\usepackage{times}
\usepackage{url}
\usepackage{graphicx}
\usepackage{boxedminipage}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{wrapfig}
\usepackage{balance}
\usepackage[protrusion=true,expansion=true]{microtype}
%\pdfminorversion=3
\newcommand{\smallurl}[1]{{\small \url{#1}}}

\newcommand{\BOOM} {BOOM\xspace}
\newcommand{\BOOMA} {BOOM Analytics\xspace}
\newcommand{\BIGBOOM} {BOOM\xspace}
\newcommand{\JOL} {JOL\xspace}
\newcommand{\BIGJOL} {JOL\xspace}
\newcommand{\JT} {JobTracker\xspace}
\newcommand{\TT} {TaskTracker\xspace}
\newcommand{\NN} {NameNode\xspace}
\newcommand{\DN} {DataNode\xspace}
\renewcommand{\ttdefault}{cmtt}

\begin{document}

\title{\BOOMA: Exploring Data-Centric, Declarative Programming for the Cloud}
\conferenceinfo{EuroSys'10,} {April 13--16, 2010, Paris, France.}
\CopyrightYear{2010}
\copyrightdata{978-1-60558-577-2/10/04}

\authorinfo{Peter Alvaro}{UC Berkeley}{palvaro@eecs.berkeley.edu}
\authorinfo{Tyson Condie}{UC Berkeley}{tcondie@eecs.berkeley.edu}
\authorinfo{Neil Conway}{UC Berkeley}{nrc@eecs.berkeley.edu}
\authorinfo{Khaled Elmeleegy}{Yahoo!\ Research}{khaled@yahoo-inc.com}
\authorinfo{Joseph M. Hellerstein}{UC Berkeley}{hellerstein@eecs.berkeley.edu}
\authorinfo{Russell Sears}{UC Berkeley}{sears@eecs.berkeley.edu}

\maketitle
\begin{abstract}
  Building and debugging distributed software remains extremely difficult. We
  conjecture that by adopting a \emph{data-centric} approach to system design
  and by employing \emph{declarative} programming languages, a broad range of
  distributed software can be recast naturally in a data-parallel programming
  model.  Our hope is that this model can significantly raise the level of
  abstraction for programmers, improving code simplicity, speed of development,
  ease of software evolution, and program correctness.

  This paper presents our experience with an initial large-scale experiment in
  this direction.  First, we used the Overlog language to implement a ``Big
  Data'' analytics stack that is API-compatible with Hadoop and HDFS and
  provides comparable performance.  Second, we extended the system with complex
  distributed features not yet available in Hadoop, including high availability,
  scalability, and unique monitoring and debugging facilities. We present both
  quantitative and anecdotal results from our experience, providing some
  concrete evidence that both data-centric design and declarative languages can
  substantially simplify distributed systems programming.
\end{abstract}

\category{H.3.4}{Information Storage and Retrieval}{Systems and
  Software---Distributed systems}

\terms Design, Experimentation, Languages

\keywords
Cloud Computing, Datalog, MapReduce

\input{intro}
\input{relwork}

\input{bg}

\input{hdfs}

\input{rely}

\input{scale}

\input{manage}

\input{mapreduce}

\input{eval}

\input{lessons}

\input{conclusion}
\input{acknowledgments}
\balance
\bibliographystyle{abbrvnat}
\bibliography{euro041-alvaro}

\end{document}
