% TEMPLATE for Usenix papers, specifically to meet requirements of
%  USENIX '05
% originally a template for producing IEEE-format articles using LaTeX.
%   written by Matthew Ward, CS Department, Worcester Polytechnic Institute.
% adapted by David Beazley for his excellent SWIG paper in Proceedings,
%   Tcl 96
% turned into a smartass generic template by De Clarke, with thanks to
%   both the above pioneers
% use at your own risk.  Complaints to /dev/null.
% make it two column with no page numbering, default is 10 point

% Munged by Fred Douglis <douglis@research.att.com> 10/97 to separate
% the .sty file from the LaTeX source template, so that people can
% more easily include the .sty file into an existing document.  Also
% changed to more closely follow the style guidelines as represented
% by the Word sample file. 
% This version uses the latex2e styles, not the very ancient 2.09 stuff.
\documentclass[letterpaper,twocolumn,10pt]{article}
\usepackage{usenix,epsfig,endnotes}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{color}
\usepackage{times}
\usepackage{url}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{subfig}

\newcommand{\JT} {JobTracker\xspace}
\newcommand{\TT} {TaskTracker\xspace}
\newcommand{\NN} {NameNode\xspace}
\newcommand{\DN} {DataNode\xspace}

% \newcommand{\paa}[1]{{\textcolor{blue}{#1 -- paa}}}
% \newcommand{\nrc}[1]{{\textcolor{green}{#1 -- nrc}}}
% \newcommand{\rcs}[1]{{\textcolor{red}{#1 -- rcs}}}
%\newcommand{\jmh}[1]{{\textcolor{red}{#1 -- jmh}}}

\begin{document}


%don't want date printed
\date{}

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
\title{\Large \bf MapReduce Online}

%for single author (just remove % characters)
\author{
{\rm Tyson Condie, Neil Conway, Peter Alvaro, Joseph M.\ Hellerstein}\\ UC Berkeley
\and
{\rm Khaled Elmeleegy, Russell Sears}\\ Yahoo! Research
}

\maketitle

\begin{abstract}
  MapReduce is a popular framework for data-intensive distributed computing of
  batch jobs. To simplify fault tolerance, many implementations of MapReduce
  \emph{materialize} the entire output of each map and reduce task before it can
  be consumed. In this paper, we propose a modified MapReduce architecture that
  allows data to be \emph{pipelined} between operators. This extends the
  MapReduce programming model beyond batch processing, and can reduce completion
  times and improve system utilization for batch jobs as well.  We present a
  modified version of the Hadoop MapReduce framework that supports \emph{online
    aggregation}, which allows users to see ``early returns'' from a job as it
  is being computed. Our Hadoop Online Prototype ({\em HOP}) also supports
  \emph{continuous queries}, which enable MapReduce programs to be written for
  applications such as event monitoring and stream processing. HOP retains the
  fault tolerance properties of Hadoop and can run unmodified user-defined
  MapReduce programs.
\end{abstract}

\input{intro}
\input{background}
\input{pipelining}
\input{onlineaggs}
\input{continuous}
\input{perf}
\input{relwork}
\input{conclusion}

\section*{Acknowledgments}
We would like to thank Daniel Abadi, Kuang Chen, Mosharaf Chowdhury, Akshay
Krishnamurthy, Andrew Pavlo, Hong Tang , and our shepherd Jeff Dean for their
helpful comments and suggestions. This material is based upon work supported by
the National Science Foundation under Grant Nos.\ 0713661, 0722077 and 0803690,
the Air Force Office of Scientific Research under Grant No.\ FA95500810352, the
Natural Sciences and Engineering Research Council of Canada, and gifts from IBM,
Microsoft, and Yahoo!.

{\footnotesize \bibliographystyle{acm}
\bibliography{nsdi}}

\end{document}
