
I would like to thank my advisor Professor Joe Hellerstein for helping me
achieve this goal.  I was first introduced to Joe by Mike Franklin, who
recommend me for a TA position in the database systems course (CS186) at U.C.\@
Berkeley.  Joe exemplified the role of a Professor and inspired me to enter
graduate school.  As a graduate student, Joe went above and beyond the call of
an advisor; working with me on developing my research, writing, and
presentation skills.  I could not have asked for a better role model.  I would
also like to thank Professors Mike Franklin, Ion Stoica, and Scott Shenker for
their guidance during my graduate career.

I spent my first two years of graduate school at the Intel Research Berkeley
Lab.  I would like to thank everyone in this lab for their assistance during
the early stages of my graduate career.  Two research scientists immediately
stand out: Petros Maniatis and Timothy Roscoe.  I will always consider Petros
my second advisor and I thank him for spending countless hours helping me
develop my research and building my confidence.  Mothy introduced me to systems
research and he is the best, and no doubt wittiest, systems programmer I have
had the privilege of working with.

I am thankful for my two summer internships at Yahoo!\@ Research; a place I now
call home.  My mentor Christopher Olston was a true inspiration in my life.  I
strive to emulate Chris' approach to research and I thank him for his guidance.
I would also like to thank Benjamin Reed, Khaled Elmeleegy, Vanja Josifovski,
Raghu Ramakrishnan, Utkarsh Srivastava, and Adam Silberstein for their
influence on my research.

I had the privilege of working with some of the best graduate students that
computer science has to offer.  Boon Thau Loo who mentored me in the P2 project
and showed me the path to a successful graduate career.  My time at Intel
Research Berkeley would not have been successful without the company of my dear
friend Atul Singh, who was always there for me in times of need.  In the latter
stages of my graduate career, I worked with Peter Alvaro and Neil Conway (the
dream team) on \BOOMA and MapReduce Online.  I would like to specifically thank
Neil Conway for his co-first authorship on the MapReduce Online paper, and
Peter Alvaro for his guidance in \BOOMA and for always saying something is good
before giving me constructive criticism.  Russell Sears also played a key role
in the development of these two projects and provided an irreplaceable presence
to the success of the \BOOM team.  I would like to thank Kuang Chen for showing
what research with real-world impact is all about and for his friendship.
Thanks to Matei Zaharia for instructing me on Hadoop.  To Alexandra Meliou for
helping me pass my preliminary examination and her leadership role in the
database group. To David Chu, Shawn Jeffery and Ryan Huebsch for helping me
develop my research topic.  And the many others that had an impact on my life
at Berkeley: Daisy Wang, Sean Rhea, Matthew Caesar, Beth Trushkowsky, Michael
Armbrust, Fred Reiss, Mehul Shah, Ashima Atul, and William Marczak.

Two experiences prior to entering graduate school at U.C.\@ Berkeley continued
to shape and motivate me.  I had the privilege of working with Professor Hector
Garcia-Molina at Stanford University.  I would like to thank Hector for
introducing me to computer science research and working with me on my first
publication.  Sep Kamvar, Mayank Bawa, and Prasanna Ganesan all played a
pivotal role in the development of my research at this early stage.  

I am also grateful for my early U.S.\@ Marine Corps experience, which motivated
me toward a higher degree --- Semper Fidelis.

Finally, I would like to dedicate this thesis to the three individuals that
made it possible.  To Professor Joe Hellerstein for navigating me through this
incredible experience and teaching me how to stand on my own two feet.  To my
uncle Paul Condie for inspiring me to enter this field and teaching me how to
think outside of the box.  To my wife Professor Lara Dolecek for leading me by
example and supporting me from the very beginning.


